Draft

### 2019-10-17 draft.tex
Formal TT, informal TT additions

\newtheorem{thm}{Theorem}[section]
\newtheorem{defn}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\crefname{thm}{Lemma}{Lemmas}

@@

\begin{titlepage}
\hypersetup{pageanchor=false}
\maketitle
\thispagestyle{empty}
\end{titlepage}
\hypersetup{pageanchor=true}
%Below is an example of the problem environment
\chapter{Type Theory}
\section{Introduction}
Martin-Löf's intensional \textit{Type Theory} is a formal language and deductive system, that is self sufficient in the sense that it need not be formulated as a collection of axioms on top of some other formal system like First Order Logic, instead its deductive system can be built on top of its own formal language.

@@

\section{Type Construction Operations}
We have the following list of types at our disposal.
\begin{itemize}
	\item Given types $A, B : \mathcal{U}$ we can define the type $A \rightarrow B$ of \textbf{functions} from $A$ to $B$. 
		We can use $\lambda$-abstraction to construct elements of this type. $\lambda x. \Phi$ lies in $A \rightarrow B$ iff for $a : A$ we have $\Phi[a/x] : B$. 
		For $f : A \rightarrow B$ and $a : A$ we have that the application of $f$ on $a$, denoted as $f(a)$ or $f\;a$, lies in $B$, so $f\;a : B$.

		Functions whose type is of the form $A \rightarrow \mathcal{U}$ i.e. the codomain is $\mathcal{U}$ are called \textit{families of types}. 
		They are of special interest because they can be viewed as types themselves; types indexed by terms of the domain type. 
		So if $B : A \rightarrow \mathcal{U}$ is such a family of types, then its inhabitants would be the collection of inhabitants of all $B(a)$'s for all $a  : A$. 
		This last sentence is not sanctioned by formal type theory and only serves to make the introduction of the first dependent type, the \textit{dependent function type}, easier.
	\item Given some type $A : \mathcal{U}$ and a family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we have the type of \textbf{dependent functions} or \textbf{dependent products} (which should not be viewed as the dependent version of the product type defined later on) $$\prod_{a : A} B(a)$$
	where for $f : \prod_{a : A} B(a)$ and $x : A$ we have $f(x) : B(x)$. 
	As in the case of non-dependent functions, we can use lambda abstraction to construct elements of a dependently-typed function type. 
	That way, $\lambda x.\Phi$ lies in $\prod_{a:A}B(a)$ iff $B$ is of the form $\lambda y.\Psi$ and for all $a:A$ we have $\Phi[a/x] : \Psi[a/y]$. 
	\item Given $A, B : \mathcal{U}$ we can define the \textbf{product} type $A \times B : \mathcal{U}$. 
	For $a : A$ and $b : B$ we have the pair $(a,b) : A \times B$. 
	We also have the projection functions 
	$$\mathtt{pr}_1 : A \times B \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : A \times B \rightarrow B : (a,b) \mapsto b$$
	We also have the following induction principle
	$$\mathtt{ind}_{A\times B} : \prod_{C: A\times B \rightarrow \mathcal{U}}\big( \prod_{a:A}\prod_{b:B} C((a,b)) \big) \rightarrow \prod_{x: A\times B} C(x)$$
	$$\mathtt{ind}_{A\times B}(C, f, (a,b)) : \equiv f(a)(b)$$
	So given two functions $f: A\rightarrow C$ and $g : B \rightarrow C$ we can construct an $h : A\times B \rightarrow C$ such that $h((a,b)) \equiv \big(f(a), g(b)\big)$ for all $a:A,\; b:B$.
	\item Given $A : \mathcal{U}$ and family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we can define the \textbf{dependent pair} type (the dependent version of the product type)
	$$\sum_{a : A} B(a)$$
	Given $x : A$ and $b : B(x)$ we can construct the pair $(x,b) : \sum_{a : A} B(a)$. 
	We have two projection functions, similar to the case of the product type.
	$$\mathtt{pr}_1 : \sum_{a:A}B(a) \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : \prod_{x : \sum_{a:A}B(a)}B(\mathtt{pr}_1\; x) : (a,b) \mapsto b$$
	The induction principle is the following
	$$\mathtt{ind}_{\sum_{a:A}B(a)} : \prod_{C : \sum_{a:A}B(a) \rightarrow \mathcal{U}}\Big( \prod_{a : A}\prod_{b : B(a)}C((a,b)) \Big)\prod_{x : \sum_{a:A}B(a)}C(x)$$
	$$\mathtt{ind}_{\sum_{a:A}B(a)}(C, f,(a,b) ) :\equiv f(a)(b)$$
	\item Given $A, B : \mathcal{U}$ we can construct the \textbf{coproduct} type $A + B$. 
	We can construct elements of $A + B$ using the functions
	$$\mathtt{inl} : A \rightarrow A+B$$
	$$\mathtt{inr} : B \rightarrow A+B$$
	One can guess the induction principle
	$$\mathtt{ind}_{A+B} : \prod_{C : A+B \rightarrow U} \Big(\prod_{a : A}C(\mathtt{inl}\;a)\Big) \rightarrow \Big(\prod_{b : B}C(\mathtt{inr}\;b)\Big) \rightarrow \prod_{x : A+ B}C(x)$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inl}\; a) :\equiv f_A\; a$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inr}\; b) :\equiv f_B\; b$$
	\item Given $x, y : A$ we have the \textbf{identity type} $x =_A y : \mathcal{U}$ (we ommit the index $A$ whenever it's easily deduced). 
	An element of this type amounts to a proof that $x$ and $y$ are equal and we call the element a path between $x$ and $y$. 
	Say $x$ and $y$ are judgmentally equal, $x \equiv y$. 
	This is captured by the element $\mathtt{idp}_x : x =_A y$. 
	The relevant induction principle describes how we can use elements of an identity type
	$$\mathtt{ind}_{=_{A}} : \prod_{C : \prod_{x,y : A} (x =_A y) \rightarrow \mathcal{U}} \Big( \prod_{x : A}C(x,x,\mathtt{idp}_x) \Big) \rightarrow \prod_{x,y : A} \prod_{(p : x =_A y)} C(x,y,p)$$
	The relevant computation gives us the judgmental (definitional) equality 
	$$\mathtt{ind}_{=_A} (C,c,x,x,\mathtt{idp_x}) \equiv c\; x$$
	We can concatenate those paths whose domains and codomains allow for it. 
	Paths form an equivalence relation. That is, if $p : x = y$ and $p' : y = z$ are such paths then we can concatenate them $p \cdot p' : x = z$. 
	We can also provide the inverse $p^{-1} : y = x$ for which we have in turn a path $q : (p\cdot p^{-1}) =_{x=x} (\mathtt{idp}_x)$ between $p\cdot p^{-1}$ and $\mathtt{idp}_x$ and another one between $p^{-1}\cdot p$ and $\mathtt{idp}_y$. 

@@

	\begin{defn}
	We call a type $A$ a \textbf{mere proposition} or simply a \textbf{proposition}, if for every $a,b : A$ we have $a = b$.
\begin{itemize}
	\item For every type $A : \mathcal{U}$ there exists its \textbf{propositional truncation} $\| A \|$. 
	We also have the truncation map $|\cdot| : A \rightarrow \lVert A \rVert$ so that for every element $a : A$ there exists $|a| : \| A\|$. 
	Finally, $\lVert A\rVert$ is a mere proposition.

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
\end{itemize}
	\begin{defn}
	Let $f,g : \prod_{a:A}B(a)$ where $B :A \rightarrow \mathcal{U}$. 
	We call a function of the following type a \textbf{homotopy} between $f$ and $g$
	$$f\sim g :\equiv \prod_{a:A}f(a) = g(a)$$
	\end{defn}

@@

\section{Homotopy Type Theory}
We get the flavour of Homotopy Type Theory (HoTT) that interests us by assuming the axiom of Univalence.
\begin{lem}
We can define the following function
$$\mathtt{idtoequiv} : (A=_\mathcal{U}B) \rightarrow (A\simeq B)$$
\end{lem}
\begin{proof}
The definition of $\mathtt{idtoequiv}$ can be found in section 2.10 of \cite{hottbook}. 
\end{proof}

\begin{axiom}[Univalence]
$\mathtt{idtoequiv}$ has contractible fibers.
\end{axiom}
In HoTT we usually assume that the universe $\mathcal{U}$ that we are working in is univalent which means that for all $A,B: \mathcal{U}$
$$(A=_\mathcal{U}B) \simeq (A \simeq B)$$
Function extensionality follows from univalence.

\section{(Definitionally) Extensional Type Theory}
Extensional Type Theory (ETT) is not consistent with the Homotopy Type Theory defined above. 
To get ETT we assume the following axiom. 
\begin{axiom}
Whenever we have an inhabitant $p : a=_C b$ we can infer $\mathtt{idp}_a : a=_C b$ i.e. $a \equiv b$. 
\end{axiom}
This axiom simplifies the landscape considerably. 
Starting with it and using induction on identity types one can eventually deduce that $p \equiv \mathtt{idp}_a$. 
Thus, the higher path structure of types collapses and types behave similarly to sets. 
Additionally, function extensionality follows from the above axiom.

This is enough talk about variants of Type Theory.
\section{Logic}
Our informal deductions in Type Theory will be reminiscent of First Order Logic ones. 
To be able to use a similar verbiage, we will set down a handful of types, corresponding to the connectives that let us form well-formed formulas in FOL. 
When working intuitionistically we might say that we have \textit{actual}/there exists \textit{actual} $a : A$ of some $A : \mathcal{U}$ 
and this constitutes a constructive proof of $A$ under the `propositions as types' regime as seen in \cite{hottbook}. 
In accordance to this and the BHK interpretation, we have the following translation of FOL connectives and quantifiers. 
Assume $A, B : \mathcal{U}$ and $C : A \rightarrow \mathcal{U}$.
\begin{itemize}
\item $A \wedge B$ is $A \times B$.
\item $A \vee B$ is $A+ B$.
\item $\forall a\in A,\; C(a)$ is simply translated to $\prod_{a : A}C(a)$. 
\item Finally, existence $ \exists a \in A,\; C(a)$ is interpreted as $\sum_{a : A}C(a)$.
\end{itemize}

We would also like a way to do classical logic in Type Theory. 
To do so we would need a way to say that some type $C$ is inhabited without providing a specific witness $c : C$, which would have the undesirable effect of providing more information than a classical statement of existence. 
An effective way to do so is to provide a witness of the truncation $\lVert C \rVert$ and say that $C$ is \textit{merely} inhabited. 
So to work classically we restrict ourselves to using only mere propositions. 
This approach is called `propositions as mere propositions' in \cite{hottbook}. 
\begin{itemize}
\item When we talk of conjunction $A \wedge B$, where $A$ and $B$ are mere propositions, we mean the product $A \times B$. 
\item We interpret $A \vee B$, where $A$ and $B$ are mere propositions, as the truncation $\| A + B\|$.
\item We interpret $\forall a\in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\prod_{a : A}P(a)$. 
\item We interpret $ \exists a \in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\| \sum_{a : A} P(a) \|$
\end{itemize}
Note that the above are all chosen so that they preserve mere propositions, e.g. $A\times B$ is a mere proposition. 
The use of the above notation is interchangeable in the sections to follow. 
We will try to specify whether a type is \textit{merely} or \textit{actually} inhabited to avoid ambiguity whenever needed.

>>>>

\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\newtheorem{exmp}[thm]{Example}
\crefname{thm}{Lemma}{Lemmas}

@@

\begin{titlepage}
\hypersetup{pageanchor=false}
\maketitle
\thispagestyle{empty}
\end{titlepage}
\hypersetup{pageanchor=true}
%Below is an example of the problem environment
\chapter{Type Theory}
\section{Introduction}
Martin-L\"of's intensional \textit{Type Theory} is a formal language and deductive system, that is self sufficient in the sense that it need not be formulated as a collection of axioms on top of some other formal system like First Order Logic, instead its deductive system can be built on top of its own formal language.

@@

\section{Type Construction Operations}\label{informalTT}
We have the following list of types at our disposal.
\begin{itemize}
	\item Given types $A, B : \mathcal{U}$ we can define the type $A \rightarrow B : \mathcal{U}$ of \textbf{functions} from $A$ to $B$. 
		We can use $\lambda$-abstraction to construct elements of this type. $\lambda x. \Phi$ lies in $A \rightarrow B$ iff for $a : A$ we have $\Phi[a/x] : B$. 
		For $f : A \rightarrow B$ and $a : A$ we have that the application of $f$ on $a$, denoted as $f(a)$ or $f\;a$, lies in $B$, so $f\;a : B$.

		Functions whose type is of the form $A \rightarrow \mathcal{U}$ i.e. the codomain is $\mathcal{U}$ are called \textit{families of types}. 
		They are of special interest because they can be viewed as types themselves; types indexed by terms of the domain type. 
		So if $B : A \rightarrow \mathcal{U}$ is such a family of types, then its inhabitants would be the collection of inhabitants of all $B(a)$'s for all $a  : A$. 
		This last sentence is not sanctioned by formal type theory and only serves to make the introduction of the first dependent type, the \textit{dependent function type}, easier.
	\item Given some type $A : \mathcal{U}$ and a family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we have the type of \textbf{dependent functions} or \textbf{dependent products} (which should not be viewed as the dependent version of the product type defined later on) $$\prod_{a : A} B(a)$$
	where for $f : \prod_{a : A} B(a)$ and $x : A$ we have $f(x) : B(x)$. 
	As in the case of non-dependent functions, we can use lambda abstraction to construct elements of a dependently-typed function type. 
	That way, $\lambda x.\Phi$ lies in $\prod_{a:A}B(a)$ iff $B$ is of the form $\lambda y.\Psi$ and for all $a:A$ we have $\Phi[a/x] : \Psi[a/y]$. 
	\item Given $A, B : \mathcal{U}$ we can define the \textbf{product} type $A \times B : \mathcal{U}$. 
	For $a : A$ and $b : B$ we have the pair $(a,b) : A \times B$. 
	We also have the projection functions 
	$$\mathtt{pr}_1 : A \times B \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : A \times B \rightarrow B : (a,b) \mapsto b$$
	We sometimes use the notation $x.\mathtt{pr}_i$ to refer to $\mathtt{pr}_i\; x$. 
	We also make use of the alias $\mathtt{fst}$ for $\mathtt{pr}_1$ and $\mathtt{snd}$ for $\mathtt{pr}_2$.

	We also have the following induction principle
	$$\mathtt{ind}_{A\times B} : \prod_{C: A\times B \rightarrow \mathcal{U}}\big( \prod_{a:A}\prod_{b:B} C((a,b)) \big) \rightarrow \prod_{x: A\times B} C(x)$$
	$$\mathtt{ind}_{A\times B}(C, f, (a,b)) : \equiv f(a)(b)$$
	So given two functions $f: A\rightarrow C$ and $g : B \rightarrow C$ we can construct an $h : A\times B \rightarrow C$ such that $h((a,b)) \equiv \big(f(a), g(b)\big)$ for all $a:A,\; b:B$.
	\item Given $A : \mathcal{U}$ and family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we can define the \textbf{dependent pair} type (the dependent version of the product type)
	$$\sum_{a : A} B(a)$$
	Given $x : A$ and $b : B(x)$ we can construct the pair $(x,b) : \sum_{a : A} B(a)$. 
	We have two projection functions, similar to the case of the product type.
	$$\mathtt{pr}_1 : \sum_{a:A}B(a) \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : \prod_{x : \sum_{a:A}B(a)}B(\mathtt{pr}_1\; x) : (a,b) \mapsto b$$
	Like in the case of the product type, we make use of the dot notation $x.\mathtt{pr}_i$ here too, along with the aliases $\mathtt{fst}$ and $\mathtt{snd}$.

	The induction principle is the following
	$$\mathtt{ind}_{\sum_{a:A}B(a)} : \prod_{C : \sum_{a:A}B(a) \rightarrow \mathcal{U}}\Big( \prod_{a : A}\prod_{b : B(a)}C((a,b)) \Big)\prod_{x : \sum_{a:A}B(a)}C(x)$$
	$$\mathtt{ind}_{\sum_{a:A}B(a)}(C, f,(a,b) ) :\equiv f(a)(b)$$
	\item Given $A, B : \mathcal{U}$ we can construct the \textbf{coproduct} type $A + B$. 
	We can construct elements of $A + B$ using the functions
	$$\mathtt{inl} : A \rightarrow A+B$$
	$$\mathtt{inr} : B \rightarrow A+B$$
	One can guess the induction principle
	$$\mathtt{ind}_{A+B} : \prod_{C : A+B \rightarrow U} \Big(\prod_{a : A}C(\mathtt{inl}\;a)\Big) \rightarrow \Big(\prod_{b : B}C(\mathtt{inr}\;b)\Big) \rightarrow \prod_{x : A+ B}C(x)$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inl}\; a) :\equiv f_A\; a$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inr}\; b) :\equiv f_B\; b$$
	\item Given $x, y : A$ we have the \textbf{identity type} $x =_A y : \mathcal{U}$ (we ommit the index $A$ whenever it's easily deduced). 
	An element of this type amounts to a proof that $x$ and $y$ are equal and we call the element a path between $x$ and $y$. 
	Say $x$ and $y$ are judgmentally equal, $x \equiv y$. 
	This is captured by the element $\mathtt{idp}_x : x =_A y$. 
	For every $x  :A$ we have $\mathtt{idp}_x : x=_A x$. 
	The relevant induction principle describes how we can use elements of an identity type
	$$\mathtt{ind}_{=_{A}} : \prod_{C : \prod_{x,y : A} (x =_A y) \rightarrow \mathcal{U}} \Big( \prod_{x : A}C(x,x,\mathtt{idp}_x) \Big) \rightarrow \prod_{x,y : A} \prod_{(p : x =_A y)} C(x,y,p)$$
	The relevant computation rule gives us the judgmental (definitional) equality 
	$$\mathtt{ind}_{=_A} (C,c,x,x,\mathtt{idp_x}) \equiv c\; x$$
	We can concatenate those paths whose domains and codomains allow for it. 
	Paths form an equivalence relation. That is, if $p : x = y$ and $p' : y = z$ are such paths then we can concatenate them $p \cdot p' : x = z$. 
	We can also provide the inverse $p^{-1} : y = x$ for which we have in turn a path $q : (p\cdot p^{-1}) =_{x=x} (\mathtt{idp}_x)$ between $p\cdot p^{-1}$ and $\mathtt{idp}_x$ and another one between $p^{-1}\cdot p$ and $\mathtt{idp}_y$.

	This would be a good place to talk a bit more about the closure properties of judgmental/definitional equality $\equiv$. 
	Essentially, we demand that $\equiv$ is a congruence relation built upon those judgmental equalities we introduce elsewhere, as part of enriching our type theory with type formers, in addition to being closed under these:
	\begin{itemize}
	\item For terms $\lambda x.\;t$ and $u$ such that the application $(\lambda x.\;t)\;u$ is legal according to their typings, we have $(\lambda x.\;t)\;u \equiv t[u/x]$. 
	\item If $t \equiv t'$ and $s \equiv s'$ and the application $t(s)$ is legal, then $t(s) \equiv t'(s')$. 
	\item Closed under lambda abstraction, i.e. if $t \equiv t'$ then $\lambda x.\;t \equiv \lambda x.\;t'$. 
	\end{itemize}
	After closing under the above, we drop from $\equiv$ all pairs of terms that belong to (judgmentally) unequal types.

	We sometimes introduce judgmental equalities with a colon on the left $:\equiv$. 
	This serves merely as an indication of `direction' in the definition wherein the left hand side is being introduced as a term while it's well understood that the right hand side is a term of type theory. 
	Sometimes the lhs is just a name, a shorthand, for the rhs. 
	The colon serves no \textit{formal} purpose and the reader may ignore it, if they do not find it helpful.

	When working in type theory, we may freely replace terms with those judgmentally equal to them.

@@

	\begin{defn}
	We call a type $A$ a \textbf{mere proposition} or simply a \textbf{proposition}, if for every $a,b : A$ we have $a = b$.
	\end{defn}
\begin{itemize}
	\item For every type $A : \mathcal{U}$ there exists its \textbf{propositional truncation} $\| A \|$. 
	We also have the truncation map $|\cdot| : A \rightarrow \lVert A \rVert$ so that for every element $a : A$ there exists $|a| : \| A\|$. 
	Finally, $\lVert A\rVert$ is a mere proposition.

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
	\item We have an \textbf{empty} type $\bot : \mathcal{U}$ with the induction principle
	$$\mathtt{ind}_\bot : \prod_{A : \mathcal{U}}\prod_{C : A \rightarrow \mathcal{U}} \prod_{x  :\bot} C\;x$$
	\item We have a unit type $\mathbf{1} : \mathcal{U}$ for which we have a term $* : \mathbf{1}$ and a witness of $\prod_{x,y : \mathbf{1}} x =y$. 
	\item We have the type of \textbf{natural numbers} $\mathbb{N} : \mathcal{U}$ equiped with $0 : \mathbb{N}$ and $\mathtt{succ} : \mathbb{N} \rightarrow \mathbb{N}$. 
	There is also the induction principle
	$$\mathtt{ind}_\mathbb{N} : \prod_{C : \mathbb{N} \rightarrow \mathcal{U}}C(0)\rightarrow \big(\prod_{ : \mathbb{N}} C(n) \rightarrow C(\mathtt{succ}\;n)\big)\rightarrow \prod_{n : \mathbb{N}}C\;n$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, 0) :\equiv f_0$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, \mathtt{succ}\;n) :\equiv f_s\big(n, \mathtt{ind}_\mathbb{N}(C,f_0, f_s, n)\big)$$

\end{itemize}
This concludes our informal presentation of the primitives of Type Theory. 
We will continue with definitions and results using the tools we laid out above. 
\begin{rem}
Sometimes, when constructing a term of some type, we say that we \textit{take cases} on it or some other type. 
By this we mean that we are invoking a type's induction or recursion principle. 
As an example, suppose we have $A, B, A+B : \mathcal{U}$ and $C : A +B \rightarrow \mathcal{U}$ and want to construct a term of $\prod_{x :A+B} C(x)$. 
We can do so by taking cases on $A+B$, i.e. provide an $f_A : \prod_{a :A}C(\mathtt{inl}\; a)$ which `describes where in $C$ we send elements of $A$' and an $f_B : \prod_{b :B}C(\mathtt{inr}\; b)$. 
We have effectively made use of the induction principle of the coproduct. 
\end{rem}
\begin{rem}
We sometimes say that we can decide whether $A+B$. 
This is just a shorthand for `we can construct a witness of $A+B$' and is inspired by the saying `we can decide P' where $P$ is a mere proposition and we mean that we can prove $P + (P \rightarrow \bot)$ which would also be a mere proposition. 
\end{rem}
\begin{rem}
Induction principles might appear to give us access only to dependent functions. 
They actually give us access to non-dependent functions too. 
Let $C : A \rightarrow \mathcal{U}$ be the family of types that forms the codomain of the resulting function of the induction principle. 
If for all $a : A$, $C(a) \equiv B$ where $B  :\mathcal{U}$ then we can form a non-dependent function $f : A \rightarrow B$. 
\end{rem}

\begin{lem}
Applying any function $f : A\rightarrow B$ to $a_1, a_2 : A$ such that $p : a_1 =_A a_2$ gives us a path $\mathtt{ap}_f (p) : f(a_1) =_B f(a_2)$. 
We name this \textbf{action on paths} and this instance, action of $f$ on $p$.\\
Formally,
$$\mathtt{ap} : \prod_{A,B : \mathcal{U}}\prod_{f : A \rightarrow B}\prod_{a_1,a_2 : A}(a_1 =_A a_2)\rightarrow \big(f(a_1) =_B f(a_2)\big)$$
\end{lem}
\begin{proof}
Assume the hypotheses $A,B,f, a_1, a_2$ and $p : a_1 =_A a_2$ as above. 
We need a witness of $f(a_1) =_B f(a_2)$. 
We use induction on the identity type $a_1 =_A a_2$. 
So we have to show that for $\mathtt{idp}_{a_1} : a_1 =_A a_2$, we have $f(a_1) =_B f(a_2)$. 
By $\mathtt{idp}_{a_1}$ we have $a_1 \equiv a_2$ which means we can rewrite the goal as $f(a_1) =_B f(a_1)$, for which we have a witness in the form of $\mathtt{idp}_{f(a_1)}$. 
\end{proof}
\section{Equivalence}
	Before going over what it means for two types to be equivalent, we have to work out some preliminary notions. 
	\begin{defn}
	Let $f,g : \prod_{a:A}B(a)$ where $B :A \rightarrow \mathcal{U}$. 
	We call a function of the following type a \textbf{homotopy} between $f$ and $g$
	$$f\sim g :\equiv \prod_{a:A}f(a) = g(a)$$
	\end{defn}

@@

\section{Univalent Type Theory}
We get the flavour of Univalent Type Theory (UTT) that interests us by assuming the axiom of Univalence.
\begin{lem}
We can define the following function
$$\mathtt{idtoequiv} : (A=_\mathcal{U}B) \rightarrow (A\simeq B)$$
\end{lem}
\begin{proof}
The definition of $\mathtt{idtoequiv}$ can be found in section 2.10 of \cite{hottbook}. 
\end{proof}

\begin{axiom}[Univalence]
$\mathtt{idtoequiv}$ has contractible fibers.
\end{axiom}
In UTT we usually assume that the universe $\mathcal{U}$ that we are working in is univalent which means that for all $A,B: \mathcal{U}$
$$(A=_\mathcal{U}B) \simeq (A \simeq B)$$
Function extensionality follows from univalence.

\section{(Definitionally) Extensional Type Theory}
Extensional Type Theory (ETT) is not consistent with the Univalent Type Theory defined above. 
To get ETT we assume the following axiom. 
\begin{axiom}
Whenever we have an inhabitant $p : a=_C b$ we can infer $\mathtt{idp}_a : a=_C b$ i.e. $a \equiv b$. 
\end{axiom}
This axiom simplifies the landscape considerably. 
Starting with it and using induction on identity types one can eventually deduce that $p \equiv \mathtt{idp}_a$. 
Thus, the higher path structure of types collapses and types behave similarly to sets. 
Additionally, function extensionality follows from the above axiom.

This is enough talk about variants of Type Theory.

\section{Logic}
Our informal deductions in Type Theory will be reminiscent of First Order Logic ones. 
To be able to use a similar verbiage, we will set down a handful of types, corresponding to the connectives that let us form well-formed formulas in FOL. 
When working intuitionistically we might say that we have \textit{actual}/there exists \textit{actual} $a : A$ of some $A : \mathcal{U}$ 
and this constitutes a constructive proof of $A$ under the `propositions as types' regime as seen in \cite{hottbook}. 
In accordance to this and the BHK interpretation, we have the following translation of FOL connectives and quantifiers. 
Assume $A, B : \mathcal{U}$ and $C : A \rightarrow \mathcal{U}$.
\begin{itemize}
\item $A \wedge B$ is $A \times B$.
\item $A \vee B$ is $A+ B$.
\item $\neg A$ is $A \rightarrow \bot$. 
\item $\forall a\in A,\; C(a)$ is simply translated to $\prod_{a : A}C(a)$. 
\item Finally, existence $ \exists a \in A,\; C(a)$ is interpreted as $\sum_{a : A}C(a)$.
\end{itemize}

We would also like a way to do classical logic in Type Theory. 
To do so we would need a way to say that some type $C$ is inhabited without providing a specific witness $c : C$, which would have the undesirable effect of providing more information than a classical statement of existence. 
An effective way to do so is to provide a witness of the truncation $\lVert C \rVert$ and say that $C$ is \textit{merely} inhabited. 
So to work classically we restrict ourselves to using only mere propositions. 
This approach is called `propositions as mere propositions' in \cite{hottbook}. 
\begin{itemize}
\item When we talk of conjunction $A \wedge B$, where $A$ and $B$ are mere propositions, we mean the product $A \times B$. 
\item We interpret $A \vee B$, where $A$ and $B$ are mere propositions, as the truncation $\| A + B\|$.
\item $\neg A$ is $A \rightarrow \bot$. 
\item We interpret $\forall a\in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\prod_{a : A}P(a)$. 
\item We interpret $ \exists a \in A,\; P(a)$, where $P(a)$ is a mere proposition for all $a \in A$, as $\| \sum_{a : A} P(a) \|$
\end{itemize}
Note that the above are all chosen so that they preserve mere propositions, e.g. $A\times B$ is a mere proposition. 
The use of the above notation is interchangeable in the sections to follow. 
We will try to specify whether a type is \textit{merely} or \textit{actually} inhabited to avoid ambiguity whenever needed. 
\section{Formal Type Theory}
We shall now have a look at what Intensional Type Theory \textit{actually is}. 
We will give a short description of a formal system in which the informal derivations we do in the rest of this text are meant to be taking place. 
Building blocks of the formal system will be juxtaposed against their informal counterparts. 
We direct those looking for a full description to Appendixes A.1 and A.2 of \cite{hottbook}.

When working informally, at any given point we have a collection of assumptions about variables and their typings that are available to be used in completing the next step of our derivation. 
This is captured by the \textit{context} which is a list of typings $x_1 : A_1,\; x_2 : A_2,\;\ldots,\;x_n : A_n$ of variables $x_i$ and their types $A_i$. 
Each $x_i$ must be distinct. 
\begin{exmp}
As one would guess
$$x_1 : \mathbf{2},\; x_2 : \mathbf{2},\; x_3 : (x_1=_\mathbf{2} x_2)$$
is a legitimate context. 
\end{exmp}

The deductive system of TT is a sort of a sequent calculus where we infer sequents called \textit{judgments} which are of the form $\Gamma \vdash a : A$ where $\Gamma$ is a context and $a : A$ the statement that $a$ is a term of type $A$. 
Both $a$ and $A$ are composed exclusively by \textit{terms over} $\Gamma$. 
By terms over $\Gamma$ we mean terms that can appear in the right hand side of valid judgments with $\Gamma$ as the left hand side, e.g. $b$ is a term over $\Gamma$ if $\Gamma \vdash b : B$ can be inferred in our system. 
The rules of inferrence prescribe how we are allowed to combine said terms and derive valid judgments.

The judgment $\Gamma \vdash a  : A$ corresponds to the exact step of an informal derivation where we construct the witness $a : A$ using terms over $\Gamma$ which we constructed in earlier steps of the same derivation. 
As one might expect, the deductive system is populated with inference rules that allow us to deduce valid judgments from collections of others; to illustrate
$$\frac{J_1,J_2,\ldots,J_n}{J}$$
All the constructions and rules described in \cref{informalTT} should and can be transcribed in this form.

There is an important judgment that subverts the form described above and which asserts the correctness of a context. 
We write it as $\Gamma\; \mathtt{ctx}$. 
There are also two rules that let us build valid contexts. 
One affirming the empty context's validity and can be the root of a deduction tree
$$\frac{~}{\cdot\; \mathtt{ctx}}$$
and one which lets us extend a given context $\Gamma$ with \textit{a type over it} (a term which, under the assumptions in $\Gamma$, can be inferred to be of type $\mathcal{U}_i$ for some $i : \mathbb{N}$)
$$\frac{\Gamma \vdash A : \mathcal{U}_i}{(\Gamma,\; x: A)\; \mathtt{ctx}}$$
where $x$ must differ from all variables in $\Gamma$. 
\begin{exmp}
Consider the case of constructing a function. 
Suppose that the current context is $\Gamma,\;x : A$ where $x$ is a variable. 
If we can deduce that $b : B$ (where $b, B$ might not be ground terms but instead contain variables like $x$ or those in $\Gamma$) then this corresponds to a valid judgment $\Gamma,\;x :A \vdash b : B$. 
Formally, this is captured by the rule
$$\frac{\Gamma,\;x : A \vdash b : B}{\Gamma \vdash \lambda(x : A).b : \prod_{x : A}B}$$
Note that the function and its type $\prod_{x : A}B$ are allowed to be dependent, as we allowed $b$ and $B$ to be open terms containing variables, that is why there is no point in specifying $B(x)$. 
\end{exmp} 
\begin{exmp}
Judgmental equalities are also taken care of by this system.
